Optimal Finite Element Integration

### 2015-08-13 TOMS_Paper/assembly.tex
Add vectorization perf eval discussion + minor fixes

\begin{figure}[h]\begin{CenteredBox}
\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_opt.code}
\end{CenteredBox}\caption{Optimized Loop nest example}\label{code:loopnest-opt}\end{figure}

@@

\begin{figure}[h]\begin{CenteredBox}
\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_inc_factor.code}
\end{CenteredBox}\caption{Optimized Loop nest example}\label{code:increase_factor}\end{figure}

@@

\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/mass}}
\caption{Performance evaluation for the mass matrix. The bars represent speed up over the original, unoptimized code produced by the FEniCS Form Compiler.}\label{fig:mass}
\end{figure}

\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/helmholtz}}
\caption{Performance evaluation for the bilinear form of a Helmholtz equation. The bars represent speed up over the original, unoptimized code produced by the FEniCS Form Compiler.}\label{fig:helmholtz}
\end{figure}

\begin{figure}
\includegraphics[scale=0.7]{perf-results/elasticity}
\caption{Performance evaluation for the bilinear form arising in a elastic model. The bars represent speed up over the original, unoptimized code produced by the FEniCS Form Compiler.}\label{fig:elasticity}
\end{figure}

\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/hyperelasticity}}
\caption{Performance evaluation for the bilinear form arising in a hyperelastic model. The bars represent speed up over the original, unoptimized code produced by the FEniCS Form Compiler.}\label{fig:hyperelasticity}
\end{figure}

@@

The first observation is that our optimality model does not imply minimum execution time. In particular, in the 0.03$\%$ of the test cases (we are not counting marginal differences), \texttt{cfO2} does not result in the best runtime. There are two reasons for this. The former is that low level optimization can have a significant impact. Compare, as an example, \texttt{tens} and \texttt{cfO2} in \texttt{Mass} and \texttt{Helmholtz} when $q=1$. With $p \in [2,3]$, \texttt{tens} slightly outperforms \texttt{cfO2}. With $p=4$, \texttt{tens} is quite faster than \texttt{cfO2}. The motivation is the aggressive unrolling performed in \texttt{tens}, which 1) allows to eliminate all operations involving zero-valued entries and 2) can drastically decrease both working set and register pressure by avoiding storing entire temporaries (there are often many repeated values that, with unrolling, need be allocated only once). In COFFEE we never unroll loops to maximize the chances of autovectorization, which usually results in considerable speed ups throughout the majority of test cases. The latter reason is simply inherent to our optimality model. As explained in Section~\ref{sec:heuristic-int-dep}, we use heuristics to optimize the integration loop. The reason we are studying an hyperelastic model is indeed to assess their effectiveness; the systematic performance improvements observed over all other variants are, however, extremely encouraging.

A second observation concerns the ``triggering'' of pre-evaluation in \texttt{cfO2}. The fact that a bar in \texttt{cfO2} matches the corresponding bar in \texttt{cfO1} usually suggests that pre-evaluation was found unprofitable. The trend -- although there are quite a few exceptions -- is that as $nf$ increases, pre-evaluation tends not to be performed. This is a consequence of a larger increase factor, as discussed in Section~\ref{sec:op_count}.

Also note some other minor things.
\begin{itemize}
\item The lack of improvements in all code variants when $[q=1, p=1]$ is due to the cost of matrix insertion, which outweighs that of local assembly. 
\item The slow downs experienced with \texttt{quad} are imputable to the presence of indirection arrays in the generated code; especially with dense tables, trading vectorization for avoiding iteration over all zero-valued columns may be counter-productive (see also Section~\ref{sec:zeros}).
\item A missing bar in \texttt{tens} or \texttt{auto} means that the code generation system failed because of either exceeding the memory limit or being unable to manipulate the math characterizing the form (this happens systematically in the hyperelastic model, and for some discretizations in the elastic model).
\item The bars for $nf=0$ do not change as $p$ increases: this is expected since varying $p$ when there are no coefficients should not make any difference.
 \end{itemize}

@@

\paragraph{Elasticity}
The performance results for the elastic model are displayed in Figure~\ref{fig:elasticity}. The fact that \texttt{auto} opts for \texttt{tens} when $nf=1$ leads, generally, to sub-optimal execution times. This is different than in the case of the Helmholtz equation, in which the choice of \texttt{tens} was generally correct when $nf=1$. Pre-evaluation is never triggered in \texttt{cfO2}, because of either being estimated sub-optimal or exceeding $T_H$. In the former case, the performance improvements exhibited by \texttt{cfO2} over \text{cfO1} (and, in general, over all code variants) are due to completely removing sharing and avoiding iteration over zero-valued blocks. The latter case occurs with $q=4$ and $nf \in [0, 1]$. Running the same experiments with $T_H = L3_{size}$ resulted in a considerable improvement when $nf = 0$, with \texttt{cfO2} becoming even faster than \texttt{tens}, while slow downs characterized the cases of $nf=1$ for all values of $p$. Our explanation for this is that when $[q=4, nf=0]$, $T_H$ is exceeded by only roughly 40$\%$, while the cost function predicts a save in operation count of 5$\times$. This suggest that our model could be refined to handle the cases in which the gain in operation count is so large to make non-extreme variations in working set size negligible.

>>>>

\begin{figure}[h]\begin{CenteredBox}
\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_opt.code}
\end{CenteredBox}\caption{A possible loop nest for the local assembly of a form after applying pre-evaluation to some of the monomials (resulting in sub-expressions \texttt{G'}, \texttt{G''}, ...)}\label{code:loopnest-opt}\end{figure}

@@

\begin{figure}[h]\begin{CenteredBox}
\lstinputlisting[basicstyle=\footnotesize\ttfamily]{listings/loopnest_inc_factor.code}
\end{CenteredBox}\caption{Simplified loop nest for the local assembly of a pre-multiplied mass matrix.}\label{code:increase_factor}\end{figure}

@@

\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/mass}}
\caption{Performance evaluation for the mass matrix. The bars represent speed up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:mass}
\end{figure}

\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/helmholtz}}
\caption{Performance evaluation for the bilinear form of a Helmholtz equation. The bars represent speed up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:helmholtz}
\end{figure}

\begin{figure}
\includegraphics[scale=0.7]{perf-results/elasticity}
\caption{Performance evaluation for the bilinear form arising in a elastic model. The bars represent speed up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:elasticity}
\end{figure}

\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/hyperelasticity}}
\caption{Performance evaluation for the bilinear form arising in a hyperelastic model. The bars represent speed up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:hyperelasticity}
\end{figure}

@@

The first observation is that our optimality model does not imply minimum execution time. In particular, in the 0.06$\%$ of the test cases (we are not counting marginal differences), \texttt{cfO2} does not result in the best runtime. There are two reasons for this. The former is that low level optimization can have a significant impact. Compare, as an example, \texttt{tens} and \texttt{cfO2} in \texttt{Mass} and \texttt{Helmholtz} when $q=1$. With $p \in [2,3]$, \texttt{tens} slightly outperforms \texttt{cfO2}. With $p=4$, \texttt{tens} is quite faster than \texttt{cfO2}. The motivation is the aggressive unrolling performed in \texttt{tens}, which 1) allows to eliminate all operations involving zero-valued entries and 2) can drastically decrease both working set and register pressure by avoiding storing entire temporaries (there are often many repeated values that, with unrolling, need be allocated only once). In COFFEE we never unroll loops to maximize the chances of autovectorization, which usually results in considerable speed ups throughout the majority of test cases. The latter reason is simply inherent to our optimality model. As explained in Section~\ref{sec:heuristic-int-dep}, we use heuristics to optimize the integration loop. The reason we are studying an hyperelastic model is indeed to assess their effectiveness; the systematic performance improvements observed over all other variants are, however, extremely encouraging.

The second aspect regards the ``triggering'' of pre-evaluation in \texttt{cfO2}. The fact that a bar in \texttt{cfO2} matches the corresponding bar in \texttt{cfO1} usually suggests that pre-evaluation was found unprofitable. The trend -- although there are quite a few exceptions -- is that as $nf$ increases, pre-evaluation tends not to be performed. This is a consequence of a larger increase factor, as discussed in Section~\ref{sec:op_count}.

The third point concerns the effect of the chosen discretizations on autovectorization. The discretizations employed ensure both multilinear loop nest and tabulated basis function sizes to be multiple of the machine vector length. Given the suitable choice of compilation options, this promotes autovectorization in the majority of code variants. The biggest exception is \texttt{quad}, in which, due to the presence of indirection arrays in the generated code, vectorization is rarely triggered. In \texttt{tens}, the loop nests are fully unrolled, so the standard loop vectorization is simply not possible; however, manual inspection of the compiled code suggests block vectorization (\cite{SLP-vect}) is often triggered. In \texttt{ufls}, \texttt{cfO1}, and \texttt{cfO2} the iteration spaces tend to have the same ``structure'' (there are very few exceptions due to low level optimization), with loop vectorization being regularly applied, as far as we could evince. It is worth noting the fact that the size of loops and tabulated basis functions is a multiple of the vector length render padding and data alignment (in \texttt{cfO1} and \texttt{cfO2}), to the best of our understanding, basically irrelevant.

Also note some other minor things.
\begin{itemize}
\item The lack of (or marginal) improvements in all code variants when $[q=1, p=1]$ is due to the cost of matrix insertion, which outweighs that of local assembly. 
\item The slow downs experienced with \texttt{quad} are imputable to the presence of indirection arrays in the generated code; especially with dense tables, trading vectorization for avoiding iteration over a few zero-valued columns may be counter-productive (see also Section~\ref{sec:zeros}).
\item A missing bar in \texttt{tens} or \texttt{auto} means that the code generation system failed because of either exceeding the memory limit or being unable to manipulate the math characterizing the form (this happens systematically in the hyperelastic model, and for some discretizations in the elastic model).
\item The bars for $nf=0$ do not change as $p$ increases: this is expected since varying $p$ when there are no coefficients should not make any difference.
 \end{itemize}

@@

\paragraph{Elasticity}
The performance results for the elastic model are displayed in Figure~\ref{fig:elasticity}. The fact that \texttt{auto} opts for \texttt{tens} when $nf=1$ leads, generally, to sub-optimal execution times. This is different than in the case of the Helmholtz equation, in which the choice of \texttt{tens} was generally correct when $nf=1$. Pre-evaluation is never triggered in \texttt{cfO2}, because of either being estimated sub-optimal or exceeding $T_H$. In the former case, the performance improvements exhibited by \texttt{cfO2} over \text{cfO1} (and, in general, over all code variants) are due to completely removing sharing and avoiding iteration over zero-valued blocks. The latter case occurs with $q=4$ and $nf \in [0, 1]$. Running the same experiments with $T_H = L3_{size}$ resulted in a considerable improvement when $nf = 0$, with \texttt{cfO2} becoming even faster than \texttt{tens}, while slow downs characterized the cases of $nf=1$ for all values of $p$. Our explanation for this is that when $[q=4, nf=0]$, $T_H$ is exceeded by only roughly 40$\%$, while the cost function predicts a save in operation count of 5$\times$. This suggests that our model could be refined to handle the cases in which the gain in operation count is so large to make non-extreme variations in working set size negligible.
